---
title: "Chimpanzees engage in competitive altruism in a triadic Ultimatum Game"
subtitle: "Model summaries"
author: "Luke Maurits <<luke_maurits@eva.mpg.de>>"
date: "`r Sys.Date()`"
output:
  pdf_document:
    number_sections: TRUE
urlcolor: blue
---

```{r setup, include=FALSE}
library(knitr)
library(matrixStats)
library(tidyverse)
library(brms)

p95 <- c(0.025, 0.975)


pretty_col_names <- c("Posterior mean",
                      "95% HPD lower",
                      "95% HPD upper")

opts_chunk$set(echo=FALSE)

set.seed(1234)
```

NB: Throughout, predictions for the "8.5th" session refer to those obtained with the centered session number predictor set to zero.  Because the total number of sessions is even, this does not directly correspond to any particular actual experimetnal session.  Nevertheless, it faithly represents expected performance "half way through" any inferred learning effects.

# Pre-registered models of total proportional offer

## M1: Proportional total offer, dyadic vs triadic

```{r}
m1 <- read_rds("model1.rds")
preds <- posterior_predict(m1,re_formula=NA,
			   newdata=expand_grid(game_type=c("dyadic", "triadic"),
					       session=0))
```

### Posterior predictive distribution for total proportional offers in "8.5th" session

```{r}
condition_values <- cbind(colMeans(preds), colQuantiles(preds, probs=p95))
colnames(condition_values) <- pretty_col_names
rownames(condition_values) <- c("Dyadic", "Triadic")
kable(condition_values, digits=2)
```

### Posterior predictive distribution for difference in total proportional offers in "8.5th" session

```{r}
contrast <- preds[,2] - preds[,1]
contrast <- c(mean(contrast), quantile(contrast, probs=p95))
names(contrast) <- pretty_col_names
contrast <- as_tibble(t(contrast))
kable(contrast, digits=2)
```

## M2: Proportional total offer, consecutive vs simultaneous

```{r}
m2 <- read_rds("model2.rds")
draws <- as_draws_df(m2)
preds <- posterior_predict(m2,re_formula=NA,
			   newdata=expand_grid(game_type=c("triadic"), consec=c(-0.5, 0.5),
					       session=0))
```
### Posterior predictive distribution for total proportional offers in "8.5th" triadic session

```{r}
consec_values <- cbind(colMeans(preds), colQuantiles(preds, probs=p95))
colnames(consec_values) <- pretty_col_names
rownames(consec_values) <- c("Simultaneous", "Consecutive")
kable(consec_values, digits=2)
```

\newpage

Posterior probability that mean consecutive offer is higher than mean simulteanous offer:

```{r}
print(mean(draws$b_consec > 0))
```

Posterior predictive distribution for difference in offer between conditions:
```{r}
contrast <- preds[,2] - preds[,1]
contrast <- c(mean(contrast), quantile(contrast, probs=p95))
names(contrast) <- pretty_col_names
contrast <- as_tibble(t(contrast))
kable(contrast, digits=2)
```

# Models of tendency for final offfers to exceed initial offers within sessions

```{r}
m_first_last <- read_rds("first_last.rds")
preds <- plogis(posterior_linpred(m_first_last, re_formula=NA,
				  newdata=expand_grid(game_type=c(-0.5, 0.5), session=1:16 - 8.5)))
pivot <- (dim(preds)[1] * dim(preds)[2]) /2
marginal_dyadic_probs <- as.vector(preds)[1:pivot]
marginal_triadic_probs <- as.vector(preds)[(pivot+1):(2*pivot)]
```

### Marginalised over session probabilities of increasing within a session

```{r}
dyadic_sum <- c(mean(marginal_dyadic_probs), quantile(marginal_dyadic_probs, probs=p95))
triadic_sum <- c(mean(marginal_triadic_probs), quantile(marginal_triadic_probs, probs=p95))
x <- rbind(dyadic_sum, triadic_sum)
colnames(x) <- pretty_col_names
rownames(x) <- c("Dyadic", "Triadic")
kable(x, digits=2)
```

### Per-session probabilities

#### Dyadic

```{r}
per_session_probs <- cbind(rep(1:16,2), colMeans(preds), colQuantiles(preds, probs=p95))
colnames(per_session_probs) <- c("Session", pretty_col_names)
kable(per_session_probs[1:16,], digits=2)
```
\newpage

#### Triadic
```{r}
kable(per_session_probs[17:32,], digits=2)
```

#### Per session contrast in probabilities (triadic - dyadic)

```{r}
contrast <- preds[,17:32] - preds[,1:16]
x <- cbind(1:16, colMeans(contrast), colQuantiles(contrast, probs=p95))
colnames(x) <-  c("Session", pretty_col_names)
kable(x, digits=2)
```

### Per-triad contrast in probabilities (traidic - dyadic) in "8.5th" session

```{r}
triads <- unique(m_first_last$data$triad_id)
nd <- expand_grid(game_type=c(-0.5, 0.5), session=0, triad_id=triads)
preds <- plogis(posterior_linpred(m_first_last, newdata=nd))
contrast <- preds[,8:14] - preds[,1:7]
x <- tibble(Triad=triads, `Posterior mean`=colMeans(contrast),
            `Post. prob. of positive contrast` = colMeans(contrast > 0))
kable(x, digits=2)
```

# Models of tendency to increase offers from trial to trial

## Relative to self

```{r}
m_self <- read_rds("increase_over_self.rds")
preds <- plogis(posterior_linpred(m_self, re_formula=NA,
			   newdata=expand_grid(game_type=c("triadic"), previous_accepted=c(-0.5, 0.5),
					       session=1:16 - 8.5)))
```

### Marginalised over sessions

```{r}
pivot <- (dim(preds)[1] * dim(preds)[2]) / 2
marginal_rejected_probs <- as.vector(preds)[1:pivot]
marginal_accepted_probs <- as.vector(preds)[(pivot+1):(2*pivot)]

accepted_sum <- c(mean(marginal_accepted_probs), quantile(marginal_accepted_probs, probs=p95))
rejected_sum <- c(mean(marginal_rejected_probs), quantile(marginal_rejected_probs, probs=p95))
x <- rbind(accepted_sum, rejected_sum)
colnames(x) <- pretty_col_names
rownames(x) <- c("Previous offer accepted", "Previous offer rejected")
kable(x, digits=2)
```

### Accepted vs rejected contrast

```{r}
contrast <- preds[,17:32] - preds[,1:16]
contrast <- c(mean(contrast), quantile(contrast, probs=p95))
names(contrast) <- pretty_col_names
kable(as_tibble(t(contrast)), digits=2)
```

### Per-proposer and responder accepted vs rejected contrasts in "8.5th" session

```{r}
nd <- m_self$data %>% select(proposer_id, responder_id) %>% unique
nd$session <- 0
nd$previous_accepted <- -0.5
rejected_preds <- plogis(posterior_linpred(m_self, newdata=nd))
nd$previous_accepted <- 0.5
accepted_preds <- plogis(posterior_linpred(m_self, newdata=nd))
contrast <- accepted_preds - rejected_preds
x <- tibble(Proposer = nd$proposer_id, Responder = nd$responder_id,
	    `Posterior mean` = colMeans(contrast),
	    `Post. prob. of negative contrast` = colMeans(contrast < 0))
kable(arrange(x, Proposer, Responder), digits=2)
```

### Separating consecutive and simultaneous trials

#### Posterior probability of higher increasing probability in consecutive trials than simultaneous

```{r}
m_consim <- read_rds("increase_over_self_consim.rds")
draws <- as_draws_df(m_consim)
mean(draws$b_type_trial > 0)
```

#### Posterior probability of increasing offers for different trial types

```{r}
sim_accepted_preds <- plogis(posterior_linpred(m_consim, re_formula=NA, newdata=expand_grid(type_trial=-0.5, previous_accepted=0.5, session=1:16 - 8.5)))
sim_rejected_preds <- plogis(posterior_linpred(m_consim, re_formula=NA, newdata=expand_grid(type_trial=-0.5, previous_accepted=-0.5, session=1:16 - 8.5)))
con_accepted_preds <- plogis(posterior_linpred(m_consim, re_formula=NA, newdata=expand_grid(type_trial=0.5, previous_accepted=0.5, session=1:16 - 8.5)))
con_rejected_preds <- plogis(posterior_linpred(m_consim, re_formula=NA, newdata=expand_grid(type_trial=0.5, previous_accepted=-0.5, session=1:16 - 8.5)))
x <- tibble(`Previous offer`=rep(c("Accepted", "Rejected"), 2),
	    `Trial type`=rep(c("Triadic simultaneous", "Triadic consecutive"), each=2),
	    `Posterior mean` = c(mean(sim_accepted_preds),
	                         mean(sim_rejected_preds),
	                         mean(con_accepted_preds),
	                         mean(con_rejected_preds)),
            `95% HPD lower` =  c(quantile(sim_accepted_preds, probs=0.025),
	                         quantile(sim_rejected_preds, probs=0.025),
	                         quantile(con_accepted_preds, probs=0.025),
	                         quantile(con_rejected_preds, probs=0.025)),
            `95% HPD upper` =  c(quantile(sim_accepted_preds, probs=0.975),
	                         quantile(sim_rejected_preds, probs=0.975),
	                         quantile(con_accepted_preds, probs=0.975),
	                         quantile(con_rejected_preds, probs=0.975)))
kable(arrange(x, `Previous offer`, `Trial type`), digits=2)
```

## Relative to previous trial winner

NB: here we consider increasing over *or* matching the previous winner.

```{r}
m_winner <- read_rds("match_winner.rds")
preds <- plogis(posterior_linpred(m_winner, re_formula=NA,
			   newdata=expand_grid(game_type=c("triadic"), previous_accepted=c(-0.5, 0.5),
					       session=1:16 - 8.5)))
```

### Marginalised over sessions

```{r}
pivot <- (dim(preds)[1] * dim(preds)[2]) / 2
marginal_rejected_probs <- as.vector(preds)[1:pivot]
marginal_accepted_probs <- as.vector(preds)[(pivot+1):(2*pivot)]

accepted_sum <- c(mean(marginal_accepted_probs), quantile(marginal_accepted_probs, probs=p95))
rejected_sum <- c(mean(marginal_rejected_probs), quantile(marginal_rejected_probs, probs=p95))
x <- rbind(accepted_sum, rejected_sum)
colnames(x) <- pretty_col_names
rownames(x) <- c("Previous offer accepted", "Previous offer rejected")
kable(x, digits=2)
```

### Accepted vs rejected contrast

```{r}
contrast <- preds[,17:32] - preds[,1:16]
contrast <- c(mean(contrast), quantile(contrast, probs=p95))
names(contrast) <- pretty_col_names
kable(as_tibble(t(contrast)), digits=2)
```

# Models of tendency of second proposers to outbid first proposers

## Without accounting for value of first offer

```{r}
m_raises <- read_rds("second_raises.rds")
draws <- as_draws_df(m_raises)
preds <- plogis(posterior_linpred(m_raises, newdata=tibble(session=1:16 - 8.5), re_formula=NA))
```

### Posterior probability that second proposer becomes more likely to outbid first proposer as sessions increase

```{r}
print(mean(draws$b_session > 0))
```

### Probability that second proposer offers more than first proposer in consecutive trials, marginalising over sessions

```{r}
marginalised_probs <- c(mean(preds), quantile(preds, probs=p95))
names(marginalised_probs) <- pretty_col_names
kable(as_tibble(t(marginalised_probs)), digits=2)
```

\newpage

### Per session probabilities that second proposer offers more than first proposer in consecutive trials

```{r}
per_session_probs <- cbind(1:16, colMeans(preds), colQuantiles(preds, probs=p95))
colnames(per_session_probs) <- c("Session", pretty_col_names)
kable(per_session_probs, digits=2)
```

### Per second proposer and responder probabilities that second proposer offers more than first proposer in consecutive trials for "8.5th" session

```{r}
nd <- m_raises$data %>% select(second_proposer_id, responder_id) %>% unique
nd$session <- 0
outbid_probs <- plogis(posterior_linpred(m_raises, newdata=nd))
nd$session <- NULL
nd$mean <- colMeans(outbid_probs)
nd$lower <- colQuantiles(outbid_probs, p=0.025)
nd$upper <- colQuantiles(outbid_probs, p=0.975)
colnames(nd) <- c("Second proposer", "Responder", pretty_col_names)
kable(arrange(nd, `Second proposer`, Responder), digits=2)
```

\newpage

## Stratifying by first offer

```{r, include=FALSE}
m_raises_by_first <- read_rds("second_raises_by_first_and_prev.rds")
draws <- as_draws_df(m_raises_by_first)
prefs <- read_csv("estimated_offer_preferences.csv")
nd <- read_csv("raising_by_first_offer_preds.csv") %>%
	select(-previously_accepted)
```

### Estimated offering preferences

A probability distribution reflecting our estimate of proposer's preferred
offers in the absence of competition was calculated on the basis of first offers
in consecutive trials.  A distribution was calculated separately for sessions
1 through 16 as follows.  In each session and for each possible offer, the
proportion of observed offers in the 1st, 2nd, 3rd,..., 8th trial equal to that
possible offer was computed.  The session probability for that offer was set
to the weighted mean of these per-trial probabilities, with each trial receiving
half the weight of the previous trial, i.e. 2nd trial data has half the weight
of 1st trial data, 3rd trial data has half the weight of 2nd trial data, and so
on.  Under this scheme, approximately 95% of the contribution to the estimate
comes from the first half of session.  The 16 session distributions are then
combined into a weighted mean where each session receives 3/4 the weight of
the previous session, i.e. influence decays across sessions less steeply than
across trials, since sessions are separated by a much longer time.  The table
below shows both the final offering probabilities and the baseline outbidding
probabilities estimated on the basis of 10,000 simulated trials of comparing
"first" and "second" offers drawn from the preferred offer distribution.

```{r}
nd %>%
  rename(offer=first_offer) %>%
  right_join(prefs, by="offer") %>%
  select(offer, probability, baseline) %>%
  unique %>%
  rename(Offer=offer, `Preference probability`=probability,
         `Baseline outbiding probability`=baseline) %>%
  kable(digits=2)
```

### Posterior probability that second proposer becomes more likely to outbid first proposer as sessions increase

#### With previous offer rejected:

```{r}
print(mean(draws$b_session >0))
```

#### With previous offer accepted:

```{r}
print(mean(draws$b_session + draws$`b_previously_accepted:session` > 0))
```

\newpage

### Outbidding probabilities in first session, stratified by first offer

```{r}
pretty_session_predictions <- function(nd, s) {
  nd %>%
  filter(session==s) %>%
  select(-session, -below_chance) %>%
  arrange(second_previously_accepted, first_offer) %>%
  rename(`First Offer`=first_offer,
         `Previously accepted`=second_previously_accepted,
         `Posterior mean`=mean,
         `95% HPD lower`=lower,
         `95% HPD upper`=upper,
         `Baseline`=baseline,
         `PP rate above baseline`=model_higher,
         `PP rate above chance`=above_chance) %>%
  return
}

kable(pretty_session_predictions(nd, 1), digits=2)
```

### Outbidding probabilities in final session, stratified by first offer

```{r}
kable(pretty_session_predictions(nd, 16), digits=2)
```

\newpage

### Per second proposer and per responder posterior predictions on outbidding initial offers of 1, 2, 3 or 4 in final session

#### Previous offer accepted

##### Posterior probabilities that outbidding rates are higher than chance

```{r}
nd_min <- m_raises_by_first$data %>% select(second_proposer_id, responder_id) %>% unique
nd <- tibble(junk=1:(4*nrow(nd_min)))
nd$session <- 7.5
nd$previously_accepted <- +0.5
nd$second_proposer_id <- rep(nd_min$second_proposer_id, 4)
nd$responder_id <- rep(nd_min$responder_id, 4)
nd$first_offer <- rep(1:4, each=nrow(nd_min))
outbid_probs <- plogis(posterior_linpred(m_raises_by_first, newdata=nd))
nd$above_chance <- colMeans(outbid_probs >= 0.5)
nd_wide <- nd %>%
	pivot_wider(id_cols=c("second_proposer_id", "responder_id"),
		    names_from=first_offer, values_from=above_chance) %>%
	rename(`Second proposer` = second_proposer_id,
               `Responder` = responder_id) %>%
	arrange(`Second proposer`, Responder)
kable(nd_wide, digits=2)
```

##### Posterior probabilities that outbidding rates are above the estimated baseline

```{r}
nd <- left_join(nd, read_csv("raising_by_first_offer_preds.csv",
			     show_col_types = FALSE) %>%
		select(first_offer, baseline) %>% unique(), by="first_offer")
nd$above_baseline <- colMeans(outbid_probs >= nd$baseline)
nd_wide <- nd %>%
	pivot_wider(id_cols=c("second_proposer_id", "responder_id"),
		    names_from=first_offer, values_from=above_baseline) %>%
	rename(`Second proposer` = second_proposer_id,
               `Responder` = responder_id) %>%
	arrange(`Second proposer`, Responder)
kable(nd_wide, digits=2)
```

\newpage

#### Previous offer rejected

##### Posterior probabilities that outbidding rates are higher than chance

```{r}
nd_min <- m_raises_by_first$data %>% select(second_proposer_id, responder_id) %>% unique
nd <- tibble(junk=1:(4*nrow(nd_min)))
nd$session <- 7.5
nd$previously_accepted <- -0.5
nd$second_proposer_id <- rep(nd_min$second_proposer_id, 4)
nd$responder_id <- rep(nd_min$responder_id, 4)
nd$first_offer <- rep(1:4, each=nrow(nd_min))
outbid_probs <- plogis(posterior_linpred(m_raises_by_first, newdata=nd))
nd$above_chance <- colMeans(outbid_probs >= 0.5)
nd_wide <- nd %>%
	pivot_wider(id_cols=c("second_proposer_id", "responder_id"),
		    names_from=first_offer, values_from=above_chance) %>%
	rename(`Second proposer` = second_proposer_id,
               `Responder` = responder_id) %>%
	arrange(`Second proposer`, Responder)
kable(nd_wide, digits=2)
```

##### Posterior probabilities that outbidding rates are above the estimated baseline

```{r}
nd <- left_join(nd, read_csv("raising_by_first_offer_preds.csv",
			     show_col_types = FALSE) %>%
		select(first_offer, baseline) %>% unique(), by="first_offer")
nd$above_baseline <- colMeans(outbid_probs >= nd$baseline)
nd_wide <- nd %>%
	pivot_wider(id_cols=c("second_proposer_id", "responder_id"),
		    names_from=first_offer, values_from=above_baseline) %>%
	rename(`Second proposer` = second_proposer_id,
               `Responder` = responder_id) %>%
	arrange(`Second proposer`, Responder)
kable(nd_wide, digits=2)
```
