---
title: "Chimpanzees engage in competitive altruism in a triadic Ultimatum Game"
subtitle: "Model summaries"
author: "Luke Maurits <<luke_maurits@eva.mpg.de>>"
date: "`r Sys.Date()`"
output:
  pdf_document:
    number_sections: TRUE
urlcolor: blue
---

```{r setup, include=FALSE}
library(knitr)
library(matrixStats)
library(tidyverse)
library(brms)

p95 <- c(0.025, 0.975)

pretty_col_names <- c("Posterior mean",
                      "95% HPD lower",
                      "95% HPD upper")

opts_chunk$set(echo=FALSE)
```

# Pre-registered models of total proportional offer

## M1: Proportional total offer, dyadic vs triadic

```{r}
m1 <- read_rds("model1.rds")
preds <- posterior_predict(m1,re_formula=NA,
			   newdata=expand_grid(game_type=c("dyadic", "triadic"),
					       session=8))
```

### Posterior predictive distribution for total proportional offers in 8th session

```{r}
condition_values <- cbind(colMeans(preds), colQuantiles(preds, probs=p95))
colnames(condition_values) <- pretty_col_names
rownames(condition_values) <- c("Dyadic", "Triadic")
kable(condition_values, digits=2)
```

### Posterior predictive distribution for difference in total proportional offers in 8th session

```{r}
contrast <- preds[,2] - preds[,1]
contrast <- c(mean(contrast), quantile(contrast, probs=p95))
names(contrast) <- pretty_col_names
contrast <- as_tibble(t(contrast))
kable(contrast, digits=2)
```

## M2: Proportional total offer, consecutive vs simultaneous

```{r}
m2 <- read_rds("model2.rds")
draws <- as_draws_df(m2)
preds <- posterior_predict(m2,re_formula=NA,
			   newdata=expand_grid(game_type=c("triadic"), consec=c(-0.5, 0.5),
					       session=8))
```
### Posterior predictive distribution for total proportional offers in 8th triadic session

```{r}
consec_values <- cbind(colMeans(preds), colQuantiles(preds, probs=p95))
colnames(consec_values) <- pretty_col_names
rownames(consec_values) <- c("Simultaneous", "Consecutive")
kable(consec_values, digits=2)
```

Posterior probability that mean consecutive offer is higher than mean simulteanous offer:

```{r}
print(mean(draws$b_consec > 0))
```

Posterior predictive distribution for difference in offer between conditions:
```{r}
contrast <- preds[,2] - preds[,1]
contrast <- c(mean(contrast), quantile(contrast, probs=p95))
names(contrast) <- pretty_col_names
contrast <- as_tibble(t(contrast))
kable(contrast, digits=2)
```
\newpage

# Models of tendency for final offfers to exceed initial offers within sessions

```{r}
m_first_last <- read_rds("first_last.rds")
preds <- plogis(posterior_linpred(m_first_last, re_formula=NA,
				  newdata=expand_grid(game_type=c("dyadic", "triadic"), session=1:16)))

marginal_dyadic_probs <- as.vector(preds)[1:64000]
marginal_triadic_probs <- as.vector(preds)[64001:128000]
```

### Marginalised over session probabilities of increasing within a session

```{r}
dyadic_sum <- c(mean(marginal_dyadic_probs), quantile(marginal_dyadic_probs, probs=p95))
triadic_sum <- c(mean(marginal_triadic_probs), quantile(marginal_triadic_probs, probs=p95))
x <- rbind(dyadic_sum, triadic_sum)
colnames(x) <- pretty_col_names
rownames(x) <- c("Dyadic", "Triadic")
kable(x, digits=2)
```

### Per-session probabilities

#### Dyadic

```{r}
per_session_probs <- cbind(rep(1:16,2), colMeans(preds), colQuantiles(preds, probs=p95))
colnames(per_session_probs) <- c("Session", pretty_col_names)
kable(per_session_probs[1:16,], digits=2)
```
#### Triadic
```{r}
kable(per_session_probs[17:32,], digits=2)
```

#### Per session contrast in probabilities (triadic - dyadic)

```{r}
contrast <- preds[,17:32] - preds[,1:16]
x <- cbind(1:16, colMeans(contrast), colQuantiles(contrast, probs=p95))
colnames(x) <-  c("Session", pretty_col_names)
kable(x, digits=2)
```

# Models of tendency to increase offers from trial to trial

## Relative to self

```{r}
m_self <- read_rds("increase_over_self.rds")
preds <- plogis(posterior_linpred(m_self, re_formula=NA,
			   newdata=expand_grid(game_type=c("triadic"), previous_accepted=c(TRUE, FALSE),
					       session=1:16)))
```

### Marginalised over sessions

```{r}
marginal_accepted_probs <- as.vector(preds)[1:64000]
marginal_rejected_probs <- as.vector(preds)[64001:128000]

accepted_sum <- c(mean(marginal_accepted_probs), quantile(marginal_accepted_probs, probs=p95))
rejected_sum <- c(mean(marginal_rejected_probs), quantile(marginal_rejected_probs, probs=p95))
x <- rbind(accepted_sum, rejected_sum)
colnames(x) <- pretty_col_names
rownames(x) <- c("Previous offer accepted", "Previous offer rejected")
kable(x, digits=2)
```

### Accepted vs rejected contrast

```{r}
contrast <- preds[,17:32] - preds[,1:16]
contrast <- c(mean(contrast), quantile(contrast, probs=p95))
names(contrast) <- pretty_col_names
kable(as_tibble(t(contrast)), digits=2)
```

## Relative to previous trial winner

```{r}
m_winner <- read_rds("increase_over_winner.rds")
preds <- plogis(posterior_linpred(m_winner, re_formula=NA,
			   newdata=expand_grid(game_type=c("triadic"), previous_accepted=c(TRUE, FALSE),
					       session=1:16)))
```

### Marginalised over sessions

```{r}
marginal_accepted_probs <- as.vector(preds)[1:64000]
marginal_rejected_probs <- as.vector(preds)[64001:128000]

accepted_sum <- c(mean(marginal_accepted_probs), quantile(marginal_accepted_probs, probs=p95))
rejected_sum <- c(mean(marginal_rejected_probs), quantile(marginal_rejected_probs, probs=p95))
x <- rbind(accepted_sum, rejected_sum)
colnames(x) <- pretty_col_names
rownames(x) <- c("Previous offer accepted", "Previous offer rejected")
kable(x, digits=2)
```

\newpage

### Accepted vs rejected contrast

```{r}
contrast <- preds[,17:32] - preds[,1:16]
contrast <- c(mean(contrast), quantile(contrast, probs=p95))
names(contrast) <- pretty_col_names
kable(as_tibble(t(contrast)), digits=2)
```

# Models of tendency of second proposers to outbid first proposers

## Without accounting for value of first offer

```{r}
m_raises <- read_rds("second_raises.rds")
draws <- as_draws_df(m_raises)
preds <- plogis(posterior_linpred(m_raises, newdata=tibble(session=1:16), re_formula=NA))
```

### Posterior probability that second proposer becomes more likely to outbid first proposer as sessions increase

```{r}
print(mean(draws$b_session >0))
```

### Probability that second proposer offers more than first proposer in consecutive trials, marginalising over sessions

```{r}
marginalised_probs <- c(mean(preds), quantile(preds, probs=p95))
names(marginalised_probs) <- pretty_col_names
kable(as_tibble(t(marginalised_probs)), digits=2)
```

### Per session probabilities that second proposer offers more than first proposer in consecutive trials

```{r}
per_session_probs <- cbind(1:16, colMeans(preds), colQuantiles(preds, probs=p95))
colnames(per_session_probs) <- c("Session", pretty_col_names)
kable(per_session_probs, digits=2)
```

\newpage

## Stratifying by first offer

```{r, include=FALSE}
m_raises_by_first <- read_rds("second_raises_by_first.rds")
nd <- read_csv("raising_by_first_offer_preds.csv")
```

### Outbidding probabilities in first session, stratified by first offer
```{r}
pretty_session_predictions <- function(nd, s) {
  nd %>%
  filter(session==s) %>%
  select(-session) %>%
  rename(`First Offer`=first_offer,
         `Posterior mean`=mean,
         `95% HPD lower`=lower,
         `95% HPD upper`=upper,
         `Baseline outbidding rate`=baseline,
         `PP rate above basline`=model_higher,
         `PP rate above chance`=above_chance) %>%
  return
}

kable(pretty_session_predictions(nd, 1), digits=2)
```

### Outbidding probabilities in final session, stratified by first offer

```{r}
kable(pretty_session_predictions(nd, 16), digits=2)
```

### Difference in outbidding probability between initial offers of 4 and 5

```{r}
nd <- expand_grid(session=16, first_offer=c(4, 5))
preds <- plogis(posterior_linpred(m_raises_by_first, newdata=nd, re_formula=NA))
contrast <- preds[,1] - preds[,2]
contrast <- c(mean(contrast), quantile(contrast, probs=p95))
names(contrast) <- pretty_col_names
contrast <- as_tibble(t(contrast))
kable(contrast, digits=2)
```
